Learning how to act: making good decisions with machine learning

### 2017-07-16 thesis/thesis_map.tex
more work on intro, adds Tiberios suggestions

\chapter*{Todo}
\begin{enumerate}
\item Add something about Granger causality - does this belong in chapter 2 or chapter 3?
\item Sell my PhD as two key contribution. 1) A technical one, the other educational and unifying (across viewpoints)
\item Add a chapter in 'PhD prose' making some statement about the intersection of prediction and causality. When is a problem causal, what bounds does predictive accuracy give on causal models? To what degree is predictive accuracy a metric on which causal models can be judged? 
\item Benchmark estimation algorithms
\item clarify causal interpretation of Dirichlet models
\item Empirical evaluation of bandit algorithms (updated to include latest suggestions). Maybe a proper software bench marking solution?
\item If comparing bandit performance, it would be interesting to include some that violate the stochastic bandit assumption. For example, if the rewards change slowly over time or jump suddenly a limited number of times, or with some small probability. 
\end{enumerate}

@@

Predicting which customers will leave or who will re-offend if granted parole also fit within the category of problems where you wish to identify a group for which a problem will occur and target some treatment to them (hospitalisation, loyalty reward, more monitoring \& support on parole, etc). For all these problems the assumptions required to treat them as pure prediction problems are;
\begin{enumerate}
\item The treatment is assumed to be effective (at least better than nothing)
\item Deciding who to treat on the based of the model predictions won't change the relationship between features and outcome 
\end{enumerate}

@@

Playing Go (and other games) is another case with some subtleties. At every turn, the AI agent has a number of actions available. The board state following each action is deterministic and given by the rules of the game. The agent can apply supervised machine learning based on millions of previous games to estimate the probability that each of these board states will lead to a win. ... this is interesting maybe come back to it ... an alternate approach would be to try to forecast the probability of a win given each action given the current board state as context ... One approach to causal inference is indeed to learn about actions from taking actions (or observing the actions that other have taken). When can we learn from the actions others have taken? When there is no confounding. And does this hold with Go? Probably because the board state encapsulates everything that should determine what move is played. Learning directly from actions and trying to generalise (can in some instances reduce the problem to standard ML)

@@

\vspace*{.3cm}
\begin{example}
\label{exm:adjusting}
Suppose we want to estimate the impact on high school graduation rates of compulsory preschool for all 4 year olds. We have a large cross-sectional data set on a group of 20 year olds that records if they attended pre-school, if they graduated high school and their parents socio-economic status (SES). We will let $X\in \set{0,1}$ indicate if an individual attended pre-school, $Y \in \set{0,1}$ indicate if they graduated high school and $Z \in \set{0,1}$ represent if they are from a low or high SES background respectively.\footnote{There has been substantial empirical work on the effectiveness of early childhood education including a landmark randomised trial, the Perry Preschool project, which ran from 1962-1967 \citep{weikart1970longitudinal}.}
\end{example}

@@

\begin{figure}
\centering
\caption{Some valid Bayesian networks for a distribution that an be factorised as $\P{X_1,X_2,X_3} = \P{X_1}\P{X_2|X_1}\P{X_3|X_2}$ (which implies $X_3 \ci X_1 | X_2$)}
\label{fig:bayesnet2}
\begin{subfigure}[t]{0.15\textwidth}
\centering
\caption{}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={observed}]
 %nodes
\node[main node](1){$X_{1}$};
\node[main node, below=of 1](2){$X_{2}$};
\node[main node, below=of 2](3){$X_{3}$};
\path[every node/.style={font=\sffamily\small}]
    (1) edge (2)
    (2) edge (3);
\end{tikzpicture}
\end{subfigure}
\begin{subfigure}[t]{0.15\textwidth}
\centering
\caption{}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={observed}]
 %nodes
\node[main node](1){$X_{1}$};
\node[main node, below=of 1](2){$X_{2}$};
\node[main node, below=of 2](3){$X_{3}$};
\path[every node/.style={font=\sffamily\small}]
    (3) edge (2)
    (2) edge (1);
\end{tikzpicture}
\end{subfigure}
\begin{subfigure}[t]{0.3\textwidth}
\centering
\caption{}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={observed}]
 %nodes
\node[main node](1){$X_{2}$};
\node[main node, below left=of 1](2){$X_{1}$};
\node[main node, below right=of 1](3){$X_{3}$};
\path[every node/.style={font=\sffamily\small}]
    (1) edge (2) edge (3);
\end{tikzpicture}
\end{subfigure}
\begin{subfigure}[t]{0.3\textwidth}
\centering
\caption{}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={observed}]
 %nodes
\node[main node](1){$X_{2}$};
\node[main node, above left=of 1](2){$X_{1}$};
\node[main node, above right=of 1](3){$X_{3}$};
\path[every node/.style={font=\sffamily\small}]
    (2) edge (1)
    (3) edge (1)
    (2) edge (3);
\end{tikzpicture}
\end{subfigure}
\end{figure}

The statement that a given graph $G$ is a Bayesian network for a distribution $P$ tells us that the distribution can be factorised over the nodes and edges in the graph. There can be no missing edges in $G$ that do not correspond to conditional independencies in $P$, (the converse is not true $G$ can have extra edges). If we let $parents_{X_{i}}$ represent the set of variables that are parents of the variable $X_{i}$ in $G$ then we can write the joint distribution as;

@@

\begin{definition}{The do-notation}
\begin{itemize}
\item $do(X=x)$ denotes an intervention that sets the random variable(s) $X$ to $x$.
\item $\P{Y|do(X)}$ is the distribution of $Y$ conditional on an \emph{intervention} that sets $X$. This notation is somewhat overloaded. It may be used represent a probability distribution/mass function or a family of distribution functions depending on whether the variables are discrete or continuous and whether or not we are treating them as fixed. For example it could represent 
\begin{itemize}
\item the probability $\P{Y=1|do(X=x)}$ as a function of $x$,
\item the probability mass function for a discrete $Y$ : $\P{Y|do(X=x)}$,
\item the probability density function for a continuous  $Y$ : $f_Y(y|do(X=x))$,
\item a family of density/mass function for $Y$ parameterised by $x$.
\end{itemize}
Where the distinction is important and not clear from context we will use one of the more specific forms above. 
\end{itemize}
\end{definition}

\begin{theorem}[Truncated product formula \citep{Pearl2000}]
\label{thm:truncated_prodcut}
If $G$ is a causal network for a distribution $P$ defined over variables $X_{1}...X_{N}$, then we can calculate the distribution after an intervention where we set $Z \subset X$ to $z$, denoted $do(Z=z)$ by dropping the terms for each of the variables in $Z$ from the factorisation given by the network.

\begin{equation}
\label{eq:truncatedproduct}
\P{X_1...X_N|do(Z=z)} =
  \begin{cases}
  \prod_{i \notin Z}\P{X_{i}|parents_{X_{i}}} & \text{if $(X_{1}...X_{N})$ consistent with $Z=z$}  \\
   0       & \text{otherwise } 
  \end{cases}
\end{equation}
\end{theorem}

Theorem \ref{thm:truncated_prodcut} does not hold for standard Bayesian networks because there are multiple valid networks for the same distribution. The truncated product formula will give different results depending on the selected network. The result is possible with causal Bayesian networks because it follows directly from the assumption that the direction of the link indicates causality. In fact, from the interventionist viewpoint of causality, the truncation product formula defines what it means for a link to be causal.

@@

Causal Bayesian networks are Bayesian networks, so results that apply to Bayesian networks carry directly across; the local Markov property states that variables are independent of their non-effects given their direct causes. Similarly the global Markov property and d-separation also hold in causal networks.

\subsubsection{Limitations of causal Bayesian networks}
A number of criticisms have been levelled at this approach to modelling causality. One is that the definition of an intervention only in terms of setting the value of one or more variables is too precise and that any real world intervention will affect many variables in complex and non-deterministic ways \citep{rickles2009causality,cartwright2007hunting}. However, by augmenting the causal graph with additional variables that model how interventions may take effect, the deterministic do operator can model more complex interventions. For example, in the drug treatment case, we assumed that all subjects complied, taking the treatment or placebo as assigned by the experimenter. But what if some people failed to take the prescribed treatment. We can model this within the framework of deterministic interventions by adding a node representing what they were prescribed (the intervention) which probabilistically influences the treatment they actually receive, see figure \ref{fig:randomized_imperfect_compliance}. Note that the fact that we no longer directly assign the treatment opens the possibility that an unobserved latent variable could affect both the actual treatment taken and the outcome.

@@

Another key issue with causal Bayesian networks is that they cannot handle cyclic dependencies between variables. Such feedback loops are common in real-life systems, for example the relationship between supply and demand in economics or predator and prey in ecology. We might regard the underlying causal mechanisms in these examples to be acyclic; the number of predators at one time influences the number of prey in the next period and so on. However, if our measurements of these variables must be aggregated over time frames that are longer than the scale at which these interactions occur the result is a cyclical dependency. Even were we able to measure on shorter timescales, we might then not have sufficient data on each variable for inference. Such problems have mostly been studied within the dynamic systems literature, typically focusing on understanding the stationary or equilibrium state of the system and making very specific assumptions about functional form in order to make problems tractable. \citet{Poole2013} compare the equilibrium approach to reasoning about cyclic problems with structural equation models, which we discuss in section \ref{sec:SEM} and which can be seen as Bayesian causal networks with additional functional assumptions.

@@

Putting aside any estimates of individual causal effects, it is possible to learn something about the distributions under treatment or placebo. Let $\cf{Y}{1}$ be a random variable representing the potential outcome if treated. The distribution of $\cf{Y}{1}$ is the distribution of $Y$ if everyone was treated. Similarly $Y^{0}$ represents the potential outcome for the placebo. The difference between the probability of recovery, across the population, if everyone was treated and the probability of recovery given placebo is $\P{\cf{Y}{1}}-\P{\cf{Y}{0}}$. We can estimate (from an experimental or observational study):
\begin{itemize}
\item $\P{Y|X=1}$, the probability that those who took the treatment will recover
\item $\P{Y|X=0}$, the probability that those who were not treated will recover
\end{itemize}

Now, for those who took the treatment, the outcome \emph{had} they taken the treatment $\cf{Y}{1}$ is the same as the observed outcome. For those who did not take the treatment, the observed outcome is the sames as the outcome \emph{had} they not taken the treatment. Equivalently stated:

@@

The assumptions $X \ci Y^{1}$ and $X \ci Y^{0}$  are referred to as ignorability assumptions \citep{Rosenbaum1983}. They state that the treatment each person receives is independent of whether they would recover if treated and if they would recover if not treated. This is justified in example \ref{exm:ranomized_experiment} due to the randomisation of treatment assignment. In general the treatment assignment will not be independent of the potential outcomes. In example \ref{exm:adjusting}, the children who attended preschool may be more likely to have graduated high school had they in fact not attended than the children who actually did not attend, $X \nci \cf{Y}{0}$. Similarly, had the poorer children who did not attend pre-school attended they might not have done as well as the children who did in fact attend, $X \nci \cf{Y}{1}$. A more general form of the ignorability assumption is to identify a set of variables $Z$ such that $X \ci Y^{1}|Z$ and $X \ci Y^{0}|Z$.

\vspace*{.3cm}
\begin{theorem}[Ignorability] If $X \ci Y^{1}|Z$ and $X \ci Y^{0}|Z$,

@@

Taking the \emph{action} $X=1$ corresponds to replacing the equation $X=f_x(\epsilon_x)$ with $X=1$. The function $f_y$ and distribution over $\epsilon_y$ does not change. This results in the interventional distribution \footnote{We have assumed the variables are discrete only for notational convenience},

@@

\begin{figure}[h]
\caption{An example of Simpson's Paradox}
\label{fig:simpsons_paradox}
\centering
\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
\node[node](1){Gender};
\node[node, below left=of 1](2){Treatment};
\node[node, below right=of 1](3){Outcome};
\path[]
	(1) edge (2) edge (3) 
	(2) edge (3);
\end{tikzpicture}
\end{figure}

@@

\begin{figure}[h]
\caption{An example of Simpson's Paradox}
\label{fig:simpsons_paradox2}
\centering
\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
\node[node](1){Blood Pressure};
\node[node, below left=of 1](2){Treatment};
\node[node, below right=of 1](3){Outcome};
\path[]
	(1) edge (3) 
	(2) edge (3)
	(2) edge (1);
\end{tikzpicture}
\end{figure}

>>>>

\chapter*{Todo}
\begin{enumerate}
\item Add something about Granger causality - does this belong in chapter 2 or chapter 3?
\item Sell my PhD as two key contribution. 1) A technical one, the other educational and unifying (across viewpoints)
\item Add a chapter in 'PhD prose' making some statement about the intersection of prediction and causality. When is a problem causal, what bounds does predictive accuracy give on causal models? To what degree is predictive accuracy a metric on which causal models can be judged? 
\item Benchmark estimation algorithms
\item clarify causal interpretation of Dirichlet models
\item Empirical evaluation of bandit algorithms (updated to include latest suggestions). Maybe a proper software bench marking solution?
\item If comparing bandit performance, it would be interesting to include some that violate the stochastic bandit assumption. For example, if the rewards change slowly over time or jump suddenly a limited number of times, or with some small probability. 
\item Does defining causality in term of the truncated product formula ensure that a causal Bayesian network is unique? It would seem no, one can ad links (although this is hardly useful) - seems like there is no reverse condition. 
\end{enumerate}

@@

Predicting which customers will leave or who will re-offend if granted parole also fit within the category of problems where you wish to identify a group for which a problem will occur and target some treatment to them (hospitalisation, loyalty reward, more monitoring \& support on parole, etc). For all these problems the assumptions required to treat them as pure prediction problems are 
\begin{enumerate}
\item The treatment is assumed to be effective (at least better than nothing)
\item Deciding who to treat on the based of the model predictions won't change the relationship between features and outcome 
\end{enumerate}

@@

Playing Go (and other games) is a case with some subtleties. At every turn, the AI agent has a number of actions available. The board state following each action is deterministic and given by the rules of the game. The agent can apply supervised machine learning based on millions of previous games to estimate the probability that each of these board states will lead to a win. ... this is interesting maybe come back to it ... an alternate approach would be to try to forecast the probability of a win given each action given the current board state as context ... One approach to causal inference is indeed to learn about actions from taking actions (or observing the actions that other have taken). When can we learn from the actions others have taken? When there is no confounding. And does this hold with Go? Probably because the board state encapsulates everything that should determine what move is played. Learning directly from actions and trying to generalise (can in some instances reduce the problem to standard ML)

@@

\vspace*{.3cm}
\begin{example}
\label{exm:adjusting}
Suppose we want to estimate the impact on high school graduation rates of compulsory preschool for all four year olds. We have a large cross-sectional data set on a group of twenty year olds that records if they attended pre-school, if they graduated high school and their parents socio-economic status (SES). We will let $X\in \set{0,1}$ indicate if an individual attended pre-school, $Y \in \set{0,1}$ indicate if they graduated high school and $Z \in \set{0,1}$ represent if they are from a low or high SES background respectively.\footnote{There has been substantial empirical work on the effectiveness of early childhood education including a landmark randomised trial, the Perry Preschool project, which ran from 1962-1967 \citep{weikart1970longitudinal}.}
\end{example}

@@

\begin{figure}
\centering
\caption{Some valid Bayesian networks for a distribution $P$ over $(X_1,X_2,X_3)$ in which $X_3$ is conditionally independent of $X_1$ given $X_2$, denoted $X_3 \ci X_1 | X_2$. Graphs (a), (b) and (c) are all a \emph{perfect map} for $P$ as the graphical stucture implies exactly the same set of independencies exhibitied by the distribution. Graph (d), like figure \ref{fig:bayesnet} does not imply any conditional independencies, and is thus a valid (but not very useful) Bayesian network representation for any distribution over three variables.}
\label{fig:bayesnet2}
\begin{subfigure}[t]{0.15\textwidth}
\centering
\caption{}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={observed}]
 %nodes
\node[main node](1){$X_{1}$};
\node[main node, below=of 1](2){$X_{2}$};
\node[main node, below=of 2](3){$X_{3}$};
\path[every node/.style={font=\sffamily\small}]
    (1) edge (2)
    (2) edge (3);
\end{tikzpicture}
\end{subfigure}
\begin{subfigure}[t]{0.15\textwidth}
\centering
\caption{}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={observed}]
 %nodes
\node[main node](1){$X_{1}$};
\node[main node, below=of 1](2){$X_{2}$};
\node[main node, below=of 2](3){$X_{3}$};
\path[every node/.style={font=\sffamily\small}]
    (3) edge (2)
    (2) edge (1);
\end{tikzpicture}
\end{subfigure}
\begin{subfigure}[t]{0.3\textwidth}
\centering
\caption{}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={observed}]
 %nodes
\node[main node](1){$X_{2}$};
\node[main node, below left=of 1](2){$X_{1}$};
\node[main node, below right=of 1](3){$X_{3}$};
\path[every node/.style={font=\sffamily\small}]
    (1) edge (2) edge (3);
\end{tikzpicture}
\end{subfigure}
\begin{subfigure}[t]{0.3\textwidth}
\centering
\caption{}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={observed}]
 %nodes
\node[main node](1){$X_{2}$};
\node[main node, above left=of 1](2){$X_{1}$};
\node[main node, above right=of 1](3){$X_{3}$};
\path[every node/.style={font=\sffamily\small}]
    (2) edge (1)
    (3) edge (1)
    (2) edge (3);
\end{tikzpicture}
\end{subfigure}
\end{figure}

The statement that a given graph $G$ is a Bayesian network for a distribution $P$ tells us that the distribution can be factorised over the nodes and edges in the graph. There can be no missing edges in $G$ that do not correspond to conditional independencies in $P$, (the converse is not true: $G$ can have extra edges). If we let $parents_{X_{i}}$ represent the set of variables that are parents of the variable $X_{i}$ in $G$ then we can write the joint distribution as;

@@

\vspace{0.5cm}
\begin{definition}{The do-notation}
\begin{itemize}
\item $do(X=x)$ denotes an intervention that sets the random variable(s) $X$ to $x$.
\item $\P{Y|do(X)}$ is the distribution of $Y$ conditional on an \emph{intervention} that sets $X$. This notation is somewhat overloaded. It may be used to represent a probability distribution/mass function or a family of distribution functions depending on whether the variables are discrete or continuous and whether or not we are treating them as fixed. For example it could represent 
\begin{itemize}
\item the probability $\P{Y=1|do(X=x)}$ as a function of $x$,
\item the probability mass function for a discrete $Y$ : $\P{Y|do(X=x)}$,
\item the probability density function for a continuous  $Y$ : $f_Y(y|do(X=x))$,
\item a family of density/mass function for $Y$ parameterised by $x$.
\end{itemize}
Where the distinction is important and not clear from context we will use one of the more specific forms above. 
\end{itemize}
\end{definition}

\vspace{0.5cm}
\begin{theorem}[Truncated product formula \citep{Pearl2000}]
\label{thm:truncated_prodcut}
If $G$ is a causal network for a distribution $P$ defined over variables $X_{1}...X_{N}$, then we can calculate the distribution after an intervention where we set $Z \subset X$ to $z$, denoted $do(Z=z)$ by dropping the terms for each of the variables in $Z$ from the factorisation given by the network. Let $\parents{X_i}$ denote the parents of the variable $X_i$ in $G$.

\begin{equation}
\label{eq:truncatedproduct}
\P{X_1...X_N|do(Z=z)} = \ind{Z = z}
  \prod_{X_i \notin Z}\P{X_{i}|\parents{X_i}}
\end{equation}
\end{theorem}

Theorem \ref{thm:truncated_prodcut} does not hold for standard Bayesian networks because there are multiple valid networks for the same distribution. The truncated product formula will give different results depending on the selected network. The result is possible with causal Bayesian networks because it follows directly from the assumption that the direction of the link indicates causality. In fact, from the interventionist viewpoint of causality, the truncated product formula defines what it means for a link to be causal.

@@

Causal Bayesian networks are Bayesian networks, so results that apply to Bayesian networks carry directly across: the local Markov property states that variables are independent of their non-effects given their direct causes. The global Markov property and d-separation also hold in causal networks. D-speration, which characterises which conditional independencies must hold in any distribution that can represented by a given Bayesian network $G$, is key to many important results and algorithms for causal inference. We include a brief introduction in section \ref{sec:dsep}.

\subsubsection{Limitations of causal Bayesian networks}
A number of criticisms have been levelled at this approach to modelling causality. One is that the definition of an intervention only in terms of setting the value of one or more variables is too precise and that any real world intervention will affect many variables in complex and non-deterministic ways \citep{rickles2009causality,cartwright2007hunting}. However, by augmenting the causal graph with additional variables that model how interventions may take effect, the deterministic do operator can model more complex interventions. For example, in the drug treatment case, we assumed that all subjects complied, taking the treatment or placebo as assigned by the experimenter. But, what if some people failed to take the prescribed treatment? We can model this within the framework of deterministic interventions by adding a node representing what they were prescribed (the intervention) which probabilistically influences the treatment they actually receive, see figure \ref{fig:randomized_imperfect_compliance}. Note that the fact that we no longer directly assign the treatment opens the possibility that an unobserved latent variable could affect both the actual treatment taken and the outcome.

@@

Another key issue with causal Bayesian networks is that they cannot handle cyclic dependencies between variables. Such feedback loops are common in real-life systems, for example the relationship between supply and demand in economics or predator and prey in ecology. We might regard the underlying causal mechanisms in these examples to be acyclic; the number of predators at one time influences the number of prey in the next period and so on. However, if our measurements of these variables must be aggregated over time frames that are longer than the scale at which these interactions occur the result is a cyclical dependency. Even were we able to measure on shorter timescales, we might then not have sufficient data on each variable for inference. Such problems have mostly been studied within the dynamical systems literature, typically focusing on understanding the stationary or equilibrium state of the system and making very specific assumptions about functional form in order to make problems tractable. \citet{Poole2013} compare the equilibrium approach to reasoning about cyclic problems with structural equation models, which we discuss in section \ref{sec:SEM} and which can be seen as Bayesian causal networks with additional functional assumptions.

@@

Putting aside any estimates of individual causal effects, it is possible to learn something about the distributions under treatment or placebo. Let $\cf{Y}{1}$ be a random variable representing the potential outcome if treated. The distribution of $\cf{Y}{1}$ is the distribution of $Y$ if everyone was treated. Similarly $Y^{0}$ represents the potential outcome for the placebo. The difference between the probability of recovery, across the population, if everyone was treated and the probability of recovery given placebo is $\P{\cf{Y}{1}}-\P{\cf{Y}{0}}$. We can estimate (from an experimental or observational study):
\begin{itemize}
\item $\P{Y=1|X=1}$, the probability that those who took the treatment will recover
\item $\P{Y=1|X=0}$, the probability that those who were \emph{not} treated will recover
\end{itemize}

Now, for those who took the treatment, the outcome \emph{had} they taken the treatment $\cf{Y}{1}$ is the same as the observed outcome. For those who did not take the treatment, the observed outcome is the same as the outcome \emph{had} they not taken the treatment. Equivalently stated:

@@

The assumptions $X \ci Y^{1}$ and $X \ci Y^{0}$  are referred to as ignorability assumptions \citep{Rosenbaum1983}. They state that the treatment each person receives is independent of whether they would recover if treated and if they would recover if not treated. This is justified in example \ref{exm:ranomized_experiment} due to the randomisation of treatment assignment. In general the treatment assignment will not be independent of the potential outcomes. In example \ref{exm:adjusting}, the children who attended preschool may be more likely to have graduated high school had they in fact not attended than the children who actually did not attend, i.e. $X \nci \cf{Y}{0}$. Similarly, had the poorer children who did not attend pre-school attended they might not have done as well as the children who did in fact attend, i.e. $X \nci \cf{Y}{1}$. A more general form of the ignorability assumption is to identify a set of variables $Z$ such that $X \ci Y^{1}|Z$ and $X \ci Y^{0}|Z$.

\vspace*{.3cm}
\begin{theorem}[Ignorability \citep{Rosenbaum1983, Pearl2000}] If $X \ci Y^{1}|Z$ and $X \ci Y^{0}|Z$,

@@

Taking the \emph{action} $X=1$ corresponds to replacing the equation $X=f_x(\epsilon_x)$ with $X=1$. The function $f_y$ and distribution over $\epsilon_y$ does not change. This results in the interventional distribution, \footnote{We have assumed the variables are discrete only for notational convenience}

@@

\begin{figure}[h]
\caption{An example of causal network that can give rise to Simpson's Paradox. In this case we should select treatment on the basis of the gender specific results.}
\label{fig:simpsons_paradox}
\centering
\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
\node[node](1){Z = Gender};
\node[node, below left=of 1](2){T = Treatment};
\node[node, below right=of 1](3){Y = Outcome};
\path[]
	(1) edge (2) edge (3) 
	(2) edge (3);
\end{tikzpicture}
\end{figure}

@@

\begin{figure}[h]
\caption{Another causal network that can exhibit Simpson's paradox. In this case, \quotes{the solution} is not to stratify on $Z$.}
\label{fig:simpsons_paradox2}
\centering
\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
\node[node](1){Z = Blood Pressure};
\node[node, below left=of 1](2){T = Treatment};
\node[node, below right=of 1](3){Y = Outcome};
\path[]
	(1) edge (3) 
	(2) edge (3)
	(2) edge (1);
\end{tikzpicture}
\end{figure}
